---
title: "Analysis of weather impact on people and the economy"
author: "Alfonso R. Reyes"
output:
  html_notebook: default
  html_document:
    keep_md: yes
---

```{r setup, include=FALSE, error=TRUE, message=FALSE}
knitr::opts_chunk$set(echo = TRUE, comment = NA, error = TRUE)
```

## Questions

1. Across the United States, which types of events (as indicated in the EVTYPE variable) are most harmful with respect to population health?

2. Across the United States, which types of events have the greatest economic consequences?

## Synopsis
<ten sentences>
weather impact on the economy: property and crops.
Does not take into account other effects such as defferred revenue, no income,
Years recorded from 1950 till 2011.
Investigate how much is the impact of the weather
Impact on the popuplation: fatalities, injured.
For the 1st question we will use the variables: EVTYPE, FATALITIES and INJURIES.
For the 2nd question, we will use the variables: EVTYPE, PROPDMG, CROPDMG, PROPDMGEXP and CROPDMGEXP.
Other analysis can later be performed such as if the weather effects have been improving or worsening, what states have suffered the largest impact, what counties have the highest economic loss, etc.



## Data Processing
The data processing consists in the following steps:

1. Download the raw data file from the internet theme(axis.text.x = element_text(angle = 30, hjust = 1))using the `download.file` function.
2. Unpack the downloaded file (535+ MB) with the function `bunzip` into a CSV file named as `dataset.csv`.
3. Load the CSV file in the object `stormdata.raw`: 902,297 observations and 37 variables.
4. Perform a quick analysis of the dataset `stormdata.raw` before applying data transformations.
5. Create a simplified dataset `stormdata` where some non-relevant variables have been removed but all observations kept. 902,297 observations and 13 variables.
6. Create a small dataset `stormdata.small` with only 12 variables and all observations still in place. The variable `BGN_DATE` converted to date type and renamed to `DATE`. The size of this file `stormdata.small.rda` is around 6 MB.
7. Clear the object `stormdata.raw` to save memory. Use only the new and smaller data frame `stormdata`
8. Create the data frames to respond the first question.
9. Create the data frame to respond to the second question.
10. Save an additional dataset to address other questions in the future.

```{r message = FALSE}
library(dplyr)
library(ggplot2)
library(gridExtra)
library(grid)
library(stringr)
library(rprojroot)
library(R.utils)
library(lubridate)

```

### Set up project folders with `rprojroot` package
```{r results='hold'}
# package rprojroot required
root <- rprojroot::is_rstudio_project
root_file <- root$make_fix_file()
# root$find_file("REAME.md")
root_file()
cat("Setting up the project folders:\n")
project.data <- root_file("data")
project.extdata <- root_file("inst/extdata")
project.R <- root_file("R")

project.data
project.extdata
project.R
```


### Downloading the data
We found a problem while trying to download the data file from the internet. The `bunzip2` format is not properly managed by the `download.file` function. So, we have to add a new function `bunzip2` from the package `R.utils` to download and unpack the compresed file. Of course, this could have been done by using the `read.csv` function to read the data file directly but that would be a "one-off" operation. To make the function `downloadZip` reusable for the future we used `bunzip2`. There is a flag in the function that allows the selection of that particular compression format.

```{r message=FALSE, warning=FALSE}
downloadZip <- function(fileUrl, outDir="./data", bzip2 = FALSE) {
  # function to download zipped and bunzip2 files and unpack
  temp <- tempfile()
  download.file(fileUrl, temp, mode = "wb")
  if (bzip2 == FALSE) {
    unzip(temp, exdir = outDir)
  }
  else {
    destfile <- paste(outDir, "dataset.csv", sep = "/")  # provide a filename
    bunzip2(temp, destname = destfile, overwrite = TRUE) # unpack bunzip2 file
  }
}
```


```{r cache=TRUE}
# URL of the dataset file
fileUrl <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"
cat("Unpacking the raw data file:\n")

outDir <- project.extdata             # folder for raw data
downloadZip(fileUrl, outDir = outDir, bzip2 = TRUE)   # download and unpack file
```


```{r cache=TRUE}
# read the CSV file to memory
dataFile <- paste(project.extdata, "dataset.csv", sep = "/")
stormdata.raw <- read.csv(dataFile)
# stormdata <- stormdata.raw          # for the moment we will use a shorter name
```

### How our raw data look?
```{r results='hold'}
# properties of the dataset
dims <- dim(stormdata.raw)
```

There are `r dims[1]` observations and `r dims[2]` variables.

The names of the variables are:
```{r}
names(stormdata.raw)
```

```{r}
as_data_frame(stormdata.raw)
```


### What variables do we keep for our analysis?

These are the variable that we consider important for the analysis:

        REFNUM, BGN_DATE, STATE, COUNTY, COUNTYNAME, EVTYPE, 
        FATALITIES, INJURIES, PROPDMG, PROPDMGEXP, CROPDMG, CROPDMGEXP, REMARKS


```{r}
stormdata <- stormdata.raw %>%
  select(REFNUM, BGN_DATE, STATE, COUNTY, COUNTYNAME, EVTYPE, 
         FATALITIES, INJURIES, PROPDMG, PROPDMGEXP, CROPDMG, CROPDMGEXP, REMARKS) %>%
  # convert BGN_DATE to date format
  mutate(DATE = mdy_hms(as.character(BGN_DATE))) %>%    # convert to date
  select(REFNUM, DATE, STATE, COUNTY, COUNTYNAME, EVTYPE, 
         FATALITIES, INJURIES, PROPDMG, PROPDMGEXP, CROPDMG, CROPDMGEXP, REMARKS) %>%
  # convert *EXP to uppercase and remove trailing spaces
  mutate(PROPDMGEXP = as.factor(toupper(str_trim(PROPDMGEXP))),      # clean the string
         CROPDMGEXP = as.factor(toupper(str_trim(CROPDMGEXP))))
  
# rm(stormdata.raw)     # release big dataset
save(stormdata, file = paste(project.data, "stormdata.rda", sep = "/"))
```

```{r}
# floods in California from Xmas 2005 to around New Year 2006
# one record improperly entered as Billions instead of millions REFNUM=605943
# get the observations for these floods in California where the damages are above 0
stormdata %>%
  filter(DATE >= "2005-12-25" & 
           DATE <= "2006-01-02" & 
           STATE == "CA" & 
           PROPDMG > 0) %>%
  arrange(DATE) %>%
  select(REFNUM, DATE, STATE, PROPDMG, PROPDMGEXP, CROPDMG, CROPDMGEXP, REMARKS)
```


Event occurring in January 2006 (id=605943) has been improperly recorded as economic losses of 150 Billion US dollars. After some research we found that the numbers are more in the 300 million dollars range. See paper USGS _____.



```{r}
# what are the states in the variable
unique(stormdata$STATE)
```


To save memory we release now the 'stormdata.raw`  object.


We have an observation ID with the variable `REFNUM`. We check if all its values are unique:

```{r results='hold'}
# REFNUM is the record id of the observation and is unique.
length(unique(stormdata$REFNUM))
range(unique(stormdata$REFNUM))
```

This is a view of the data frame that we will use in our analysis.
```{r}
as_data_frame(stormdata)
```

### Clean up the dataset
We will transform the variable `BGN_DATE` from a factor to a date variable. We will later use it for our summaries. We will also take out some variables that are not relevant to our study.

```{r}
stormdata.small <- stormdata %>% 
  select(REFNUM, BGN_DATE, STATE, COUNTY, COUNTYNAME, EVTYPE, # take out
           FATALITIES, INJURIES, PROPDMG, PROPDMGEXP, CROPDMG, CROPDMGEXP) %>%
  mutate(DATE = mdy_hms(as.character(BGN_DATE))) %>%    # convert to date
  select(REFNUM, DATE, STATE, COUNTY, COUNTYNAME, EVTYPE,     # reorder
         FATALITIES, INJURIES, PROPDMG, PROPDMGEXP, CROPDMG, CROPDMGEXP) 

as_data_frame(stormdata.small)
```

## Other questions
### what is the worst weather event in 2005?
```{r}
# what is the worst weather event in 2005?
worst <- stormdata.small %>%
  select(REFNUM, DATE, STATE, EVTYPE, PROPDMG, PROPDMGEXP) %>%
  mutate(PROPDMG = ifelse(toupper(PROPDMGEXP) == "M", PROPDMG/1000, 
                          ifelse(toupper(PROPDMGEXP) == "K", PROPDMG/1E6,
                                 ifelse(toupper(PROPDMGEXP) == "B", PROPDMG, 0)))) %>%
  #filter(PROPDMGEXP == "B") %>%
  arrange(desc(PROPDMG))

worst
```

```{r}
events <- stormdata %>%
  select(REFNUM, REMARKS, EVTYPE) %>%
  filter(REFNUM %in% c(605943, 577616, 577615, 581535))
events
```


### Saving a portion of the dataset
If we save the data frame `stormdata` as an .rda file the size is 46 megabytes. On the other hand, if we omit the `REMARKS` variable, the new dataset shrinks to only 4.6 megabytes. We will take this route of saving the smaller file.

```{r}
# save unique events by year
save(stormdata.small, file = paste(project.data, "stormdata.small.rda", sep = "/"))
```


### The recorded weather events
Event Types `EVTYPE`

```{r}
# want to know how many levels this factor has
as_data_frame(unique(stormdata$EVTYPE), 10)

```

There are `r length(unique(stormdata$EVTYPE))` different type of events.

### Create data frames for 1st question
We want to find now which type of events is more harmful to population health. We could group by `EVTYPE` and showing the variables FATALITIES and INJURIES.

```{r}
# byEvent.0: dataset where main sort criteria is FATALITIES
byEvent.0 <- stormdata %>%
  select(EVTYPE, FATALITIES, INJURIES) %>%
  group_by(EVTYPE) %>%
  summarize(fatal.sum = sum(FATALITIES), injur.sum = sum(INJURIES)) %>%
  arrange(desc(fatal.sum), desc(injur.sum))

byEvent.0
```

```{r}
# byEvent.1: dataset where the main sort criteria is INJURIES
byEvent.1 <- stormdata %>%
  select(EVTYPE, FATALITIES, INJURIES) %>%
  group_by(EVTYPE) %>%
  summarize(injur.sum = sum(INJURIES), fatal.sum = sum(FATALITIES)) %>%
  arrange(desc(injur.sum), desc(fatal.sum))

byEvent.1
```

     
We plot now the top 5 events that cause more harm on the population:

```{r fig.asp=1}

byEvent.005 <- byEvent.0[1:5, ]

p1 <- ggplot(byEvent.005, aes(x = reorder(EVTYPE, -fatal.sum), y = fatal.sum)) +
  geom_bar(stat = "identity") +
  xlab("Event Type") + ylab("Fatalities") +
  geom_text(aes(label=fatal.sum, vjust = -0.25))

p2 <- ggplot(byEvent.005, aes(x = reorder(EVTYPE, -injur.sum), y = injur.sum)) +
  geom_bar(stat = "identity") +
  xlab("Event Type") + ylab("Injuries") +
  geom_text(aes(label=injur.sum, vjust = -0.25))

gridExtra::grid.arrange(p1, p2)
grid.rect(gp=gpar(fill=NA))
```

Tornados, Excessive heat, flash floods, heat and lightning are the weather events most harmful to the population accross the United States.

## Assessing the Economic Damage
The property and crop damage are not in a unique monetary units; they use thousands, millions and billions. They are specified in the variables `PROPDMGEXP` and `CROPDMGEXP`.

We will start by converting the monetary damages to a consistent units. We will choose thousands.

```{r}
byDamage <- stormdata %>%
  select(EVTYPE, PROPDMG, PROPDMGEXP, CROPDMG, CROPDMGEXP) %>%
  group_by(EVTYPE) %>%
  mutate(PROPDMGEXP = as.factor(toupper(str_trim(PROPDMGEXP))),      # clean the string
         CROPDMGEXP = as.factor(toupper(str_trim(CROPDMGEXP)))) %>%
  mutate(PROPDMG.K = ifelse(PROPDMGEXP == "K", PROPDMG * 1,
                     ifelse(PROPDMGEXP == "M", PROPDMG * 1000,
                     ifelse(PROPDMGEXP == "B", PROPDMG * 1E6, 0)))) %>%
  mutate(CROPDMG.K = ifelse(CROPDMGEXP == "K", CROPDMG * 1,
                     ifelse(CROPDMGEXP == "M", CROPDMG * 1000,
                     ifelse(CROPDMGEXP == "B", CROPDMG * 1E6, 0))))
byDamage
```

We convert the thousands to millions of US$ and only one variable, the total economic damage.

```{r}
byDamage.million <- byDamage %>%
  select(EVTYPE, PROPDMG.K, CROPDMG.K) %>%
  mutate(propdmg.m = PROPDMG.K /1000, cropdmg.m = CROPDMG.K / 1000) %>%
  summarize(propdmg = sum(propdmg.m), cropdmg = sum(cropdmg.m)) %>%
  mutate(damage.total = propdmg + cropdmg) %>%
  arrange(desc(damage.total))

byDamage.million
```


```{r}

byDamage.mm <- byDamage %>%
  summarize(propdmg.k = sum(PROPDMG.K), cropdmg.k = sum(CROPDMG.K)) %>%
  mutate(propdmg.m = propdmg.k / 1000, cropdmg.m = cropdmg.k / 1000) %>%
  select(EVTYPE, propdmg.m, cropdmg.m) %>%
  mutate(totaldmg.mm = propdmg.m + cropdmg.m) %>%
  mutate(totaldmg.bi = totaldmg.mm / 1000) %>%
  # arrange(desc(propdmg.m), desc(cropdmg.m))
  arrange(desc(totaldmg.mm))

byDamage.mm
```

Get the top 5 and top 10 causes of economic damage.
```{r}
byDamage.mm.top5 <- byDamage.mm[1:10, ]
byDamage.mm.top5
```

```{r fig.asp=1}
# plots for economic losses
r1 <- ggplot(byDamage.mm.top5, aes(x = reorder(EVTYPE, -totaldmg.bi), y = totaldmg.bi)) +
  geom_bar(stat = "identity") +
  labs(y = "Billions US$", x = "Weather event") +
  ggtitle("Total impact on Economy") +
  geom_text(aes(label=round(totaldmg.bi, 0), vjust = -0.25)) +
  theme(axis.text.x = element_text(angle = 30, hjust = 1))

propdmg.bi <- byDamage.mm.top5$propdmg.m/1000

r2 <- ggplot(byDamage.mm.top5, aes(x = reorder(EVTYPE, -propdmg.bi), y = propdmg.bi)) +
  geom_bar(stat = "identity") +
  labs(y = "Billions US$", x = "Weather event") +
  ggtitle("Economic impact on Property") +
  geom_text(aes(label=round(propdmg.bi, 0), vjust = -0.25)) + 
theme(axis.text.x = element_text(angle = 30, hjust = 1))

cropdmg.bi <- byDamage.mm.top5$cropdmg.m/1000

r3 <- ggplot(byDamage.mm.top5, aes(x = reorder(EVTYPE, -cropdmg.bi), y = cropdmg.bi)) +
  geom_bar(stat = "identity") +
  labs(y = "Billions US$", x = "Weather event") +
  ggtitle("Economic impact on Crops") +
  geom_text(aes(label=round(cropdmg.bi, 1), vjust = -0.25)) +
  theme(axis.text.x = element_text(angle = 30, hjust = 1))

gridExtra::grid.arrange(r1, arrangeGrob(r2, r3), ncol=2)
grid.rect(gp=gpar(fill=NA))
```


### Multiple identifiers for monetary units
There are some unspecified units in `PROPDMGEXP` and `CROPDMGEXP`.
There is no a reasonable way to determine the units or damage value from the remarks. Sometimes is thousands or in 10K, or other. Besides the identifiers `B`, `M` and `K`, there are additional characters and numbers entered in this variable. Since there is no way to etermine the units for the property or crop damage we are not considering these amounts. In two cases, we found that instead of "M" for millions the lowercase version of it "m" was used. We converted them to uppercase before summarizing the data.

Other characters or digits did not bring a special meaning to the dollar amount, so we didn't convert them even thoiugh we read the remarks to find some relationship.


```{r}
unique(byDamage$PROPDMGEXP)
```

```{r}
unique(byDamage$CROPDMGEXP)
```

```{r}
summary(byDamage)

```


Here is how we know how many observations do not carry the "K", "M" and "B" identifiers.

```{r}
unknown <- stormdata %>%
  select(STATE, EVTYPE, PROPDMGEXP, CROPDMGEXP, REMARKS) %>%
  filter(!toupper(PROPDMGEXP) %in% c("K", "M", "B") | !toupper(CROPDMGEXP) %in% c("K", "M", "B"))

```

There are `r dim(unknown)[1]` observations which dollar amount units are not properly identified in `PROPDMGEXP` and `CROPDMGEXP` variables.


### Preparing the data for other questions
We save few datasets that are much smaller in size than the original dataset.

1. byYearEvent: is a dataset that contains a summary of the events, fatalities, injuries, economic losses in property and crops in millions of US$. TO-DO: convert dates to year.

2. byYearSummary: a dataset showing the year and the wather impact on life and the economy. Four variables: year, fatalities, injured and economy losses in millions of USD.

```{r}
# By year, by event
byYearEvent <- stormdata %>%
  group_by(BGN_DATE, EVTYPE) %>%
  mutate(PROPDMGEXP = as.factor(toupper(str_trim(PROPDMGEXP))), 
         CROPDMGEXP = as.factor(toupper(str_trim(CROPDMGEXP))))     %>%
  mutate(PROPDMG.K = ifelse(PROPDMGEXP == "K", PROPDMG * 1,
                     ifelse(PROPDMGEXP == "M", PROPDMG * 1000,
                     ifelse(PROPDMGEXP == "B", PROPDMG * 1E6, 0)))) %>%
  mutate(CROPDMG.K = ifelse(CROPDMGEXP == "K", CROPDMG * 1,
                     ifelse(CROPDMGEXP == "M", CROPDMG * 1000,
                     ifelse(CROPDMGEXP == "B", CROPDMG * 1E6, 0)))) %>%
  summarize(fatalities = sum(FATALITIES), 
            injuries   = sum(INJURIES),
            propdmg.M  = sum(PROPDMG.K) / 1000,
            cropdmg.M  = sum(CROPDMG.K) /1000
            )

byYearEvent
```


```{r}
# save unique events by year
save(byYearEvent, file = paste(project.data, "byYearEvent.rda", sep = "/"))
```

### Have the number of fatalities and injuries increased over the years?

```{r}
#load(paste(project.data, "byYearEvent.rda", sep = "/"))  # load the data

byYearSummary <- byYearEvent %>%
  mutate(year = year(mdy_hms(as.character(BGN_DATE)))) %>%
  group_by(year) %>%
  summarize(fatalities = sum(fatalities), 
            injuries = sum(injuries),
            damage.mm = sum(propdmg.M) + sum(cropdmg.M),
            damage.bb = damage.mm / 1000) %>%
  arrange(desc(damage.bb))

byYearSummary
```

There are two years with the most economic damage: 2006 and 2005.


```{r fig.asp=1}
q1 <- ggplot(byYearSummary, aes(x = year, y = damage.bb)) +
  geom_point() + 
  ggtitle("Impact on economy 1950-2011") + 
  labs(y = "Billions US$") +
  theme(plot.title = element_text(hjust=0.5))
q2 <- ggplot(byYearSummary, aes(x = year, y = fatalities)) +
  geom_point() +
  ggtitle("impact on human life, 1950-2011") +
  labs(y = "Fatalities") +
    theme(plot.title = element_text(hjust=0.5))

q3 <- ggplot(byYearSummary, aes(x = year, y = injuries)) +
  geom_point() +
  ggtitle("impact on health, 1950-2011") +
  labs(y = "Injured") +
    theme(plot.title = element_text(hjust=0.5))
  

gridExtra::grid.arrange(q1, arrangeGrob(q2, q3), ncol=2)
grid.rect(gp=gpar(fill=NA))
```



```{r}
# save losses by year
save(byYearSummary, file = paste(project.data, "byYearSummary.rda", sep = "/"))
```

## Results
<present the results here>
These are the results:

1. Tornados, Excessive heat, flash floods, heat and lightning are the weather events most harmful to the population accross the United States.

2. In the case of injuries caused to humans tornados, excessive heat, lightning, heat and flash floods are the events causing them in descending order.

3. On the economic impact, we classified the weather events from the most damaging to the economy to the less. They are: flood, hurricane/typhoon, tornados, storm surge, hail, flash flodd, drought, river flood and ice storms. The economic impact totals both, property and crops.

4. Other additional findings were that the weather events have been increasing its damage to the economy and for the population in general. From the plots, we can see that from 1992 onwards, the effects have been disastrous. The losses in the economic have ascended from few millions to tens or hundreds of billions! On the population we can appreciate an increase in mortality from about a hundred in average until the end of the 80s to five hundred to fifteen hundred fatalities. 

5. Injuries have had less devastating effect but has not duplicated or triplicated as in the economy or mortality. Additional studies may be advisable to find the reason to the control in injuries in humans because of the weather events. We can notice some cyclical shape on the injuries accross the years but the slope is mildly ascending.

6. There are a couple of major weather events in 2005 and 2006 where we can see a drastic impact on the economy of 100 and 125 billion dollars. One was Katrina affecting several states in the Southm, and the other was a major flood in California.


## Figures
Maximum: 03. Can use panels.

## Code


